Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2024-10-14 paper.tex
Updates from reviews

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
steady solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We further examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs only for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The combined analytical and numerical framework presented here allows
quantifying the importance of various terms in the full wave conservation
equations.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx 0.2$.
As a corollary, except for a subset of initial conditions, the solutions to
the fully nonlinear crest-action conservation equations are mostly steady in
the reference frame of the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$, and $\psi$ is the long-wave phase.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a nonlinear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian crest and action
conservation, Hamiltonian, Schr\" odinger) that relatively closely reproduce
steady solutions for short-wave modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

@@

In this paper, we revisit this problem from linear wave theory but consider
the long-wave slope to be significant enough to require evaluating the
short-waves at the long-wave surface rather than at the mean water level.
This allows us to derive alternative, steady, nonlinear solutions for the
modulation of short waves.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

@@

Short waves that ride on the surface of the longer waves move in an accelerated
reference frame (due to the centripetal acceleration at the long-wave surface)
and thus experience effective gravitational acceleration that varies in space
and time \citep{phillips1981dispersion,longuet1986eulerian,longuet1987propagation}.
Combine (\ref{eq:dispersion}) and (\ref{eq:wave_crests}) to get:

@@

\begin{equation}
\label{eq:phi_L}
\phi_L = \dfrac{a_L \omega_L}{k_L} e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
so, up to the third order in $\varepsilon_L$, the long-wave induced horizontal
and vertical orbital velocities, evaluated at the wave surface $z = \eta_L$, are:

\begin{equation}
\label{eq:U_L}
U = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \cos{\psi}
\end{equation}

\begin{equation}
\label{eq:w_L}
W = a_L \omega_L e^{\varepsilon_L \cos{\psi}} \sin{\psi}
\end{equation}
Evaluating these velocities at the wave surface $z = \eta_L$ rather than at
the mean water level $z=0$ is a departure from the linear theory, making the
treatment of the long waves here as weakly nonlinear.

@@

\begin{equation}
\label{eq:eta_stokes}
\eta_{St} = a_L \left[
  \cos{\psi} +
  \dfrac{1}{2} \varepsilon_L \cos{2\psi} +
  \varepsilon_L^2 \left( \dfrac{3}{8} \cos{3\psi} - \dfrac{1}{16} \cos{\psi} \right)
\right]
\end{equation}
Without considering the curvilinear effects in (\ref{eq:gravity_modulation_general}),
the gravitational acceleration at the surface of a Stokes wave is, to the third
order:

@@

\begin{equation}
\label{eq:wave_amplitude_modulation}
\dfrac{\widetilde{a}}{a} = \sqrt{
  \dfrac{g}{\widetilde{g}}
  \dfrac{\widetilde{\sigma}}{\sigma}
  \dfrac{\widetilde{N}}{N}}
=
  \left( \dfrac{g}{\widetilde{g}} \right)^{0.25}
  \left( \dfrac{\widetilde{k}}{k} \right)^{0.25}
  \left( \dfrac{\widetilde{N}}{N} \right)^{0.5}
\end{equation}
To the first order in $\varepsilon_L$, and using (\ref{eq:k_short_exact})
and (\ref{eq:wave_action_modulation}), the above simplifies to:

\begin{equation}
\label{eq:wave_amplitude_modulation_order1}
\dfrac{\widetilde{a}}{a} = 
  \left( \dfrac{g}{\widetilde{g}} \right)^{0.25}
  \left( 1 + \varepsilon_L \right)^{0.75}
\end{equation}

@@

To quantify the contribution of the nonlinear terms in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}), and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are discretised using
second-order central finite difference in space and integrated in time using the
fourth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 100 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, thus neglecting the Stokes drift
induced by long waves on short-wave groups
\citep{stokes1847,van2018stokes,monismith2020stokes}.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the more sophisticated MUSCL4 scheme \citep{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centred finite difference in this model
is $\approx 0.066\%$.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

@@

\backsection[Acknowledgements]{
Thanks to Nathan Laxague, Fabrice Ardhuin, Nick Pizzo, and Peisen Tan for the
helpful discussions.
}

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear (Airy) wave theory and derive higher-order
steady solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We further examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs only for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The combined analytical and numerical framework presented here allows
quantifying the importance of various terms in the full wave conservation
equations.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx 0.2$.
As a corollary, except for a subset of initial conditions, the solutions to
the fully nonlinear crest-action conservation equations are mostly steady in
the reference frame of the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L \cos{\psi})
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$, and $\psi$ is the long-wave phase.
This process is illustrated in Fig. \ref{fig:hydrodynamic_modulation_diagram}.
Hydrodynamic modulation is a key process for the development and
interpretation of remote sensing of ocean surface waves and currents
\citep{keller1975microwave,hara1994hydrodynamic}, and is distinguished
from aerodynamic modulation \citep{donelan1987effect,belcher1999wave,chen2000effects}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited the
hydrodynamic modulation problem by considering nonlinear long waves and
variations in the effective gravity of short waves.
Their results yielded significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves using Hamiltonian mechanics and reported similar modulation magnitudes
to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a nonlinear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian crest and action
conservation, Hamiltonian, Schr\" odinger) that tackle the same fundamental
physics and that relatively closely reproduce steady solutions for short-wave
modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem remain of interest, especially if higher degrees of nonlinearity can be
considered using simpler approaches.

@@

In this paper, we revisit this problem from the perspective of linear wave theory
but consider the long-wave slope to be significant enough to require evaluating
the short-waves at the long-wave surface rather than at the mean water level.
This allows us to derive alternative, steady, nonlinear solutions for the
modulation of short waves, within the limits of small $\varepsilon_L$.
The analytical solutions yield modulations that are stronger than the original
solution by \citet{longuet1960changes}, but similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution} in
long-wave steepness up to $\approx 0.2$.
To validate the steadiness of the analytical solutions, we perform numerical
simulations of the full wave crest and action conservation laws, and compare
the results to the analytical solutions presented here, as well as the prior
numerical results.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarised in Table
\ref{table:modulation_literature_summary}.
We first present the governing equations in \S\ref{section:governing_equations},
and then the analytical solutions for the modulation of short waves by long waves
in \S\ref{section:analytical_solutions}.
Then, we describe the numerical simulations and their results in
\S\ref{section:numerical_solutions}.
Finally, we discuss the results and their implications in
\S\ref{section:discussion} and conclude the paper in \S\ref{section:conclusions}.

@@

Short waves that ride on the surface of the longer waves move in an accelerated
reference frame (due to the centripetal acceleration at the long-wave surface)
and thus experience effective gravitational acceleration that varies in space
and time \citep{phillips1981dispersion,longuet1986eulerian,longuet1987propagation}.
Inserting (\ref{eq:dispersion}) into (\ref{eq:wave_crests}) and noting that
$\partial \sqrt{gk} / \partial x = 1/2 \sqrt{g/k} \partial k / \partial x + 1/2 \sqrt{k/g} \partial g / \partial x$,
we get:

@@

\begin{equation}
\label{eq:phi_L}
\phi_L = \dfrac{a_L \omega_L}{k_L} e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
where $z$ is the vertical distance from the mean water level that is positive
upwards.
(\ref{eq:phi_L}) is accurate to the third order in $\varepsilon_L$.
The long-wave induced horizontal and vertical orbital velocities, evaluated at
the wave surface $z = \eta_L$, are:

\begin{equation}
\label{eq:U_L}
U = a_L \omega_L e^{k_L z} \cos{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}

\begin{equation}
\label{eq:w_L}
W = a_L \omega_L e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
Evaluating these velocities at the wave surface $z = \eta_L$ rather than at
the mean water level $z=0$ is a departure from the linear theory, making the
treatment of the long waves here as weakly nonlinear.
As \citet{zhang1990evolution} point out, this is appropriate to do when the
amplitude of the long waves is of the similar magnitude or large than the
wavelength of the short waves.

@@

\begin{equation}
\label{eq:eta_stokes}
\eta_{St} = a_L \left[
  \cos{\psi} +
  \dfrac{1}{2} \varepsilon_L \cos{2\psi} +
  \varepsilon_L^2 \left( \dfrac{3}{8} \cos{3\psi} - \dfrac{1}{16} \cos{\psi} \right)
\right] + \mathcal{O}(\varepsilon_L^4)
\end{equation}
Without considering the curvilinear effects in (\ref{eq:gravity_modulation_general}),
the gravitational acceleration at the surface of a Stokes wave is, to the third
order:

@@

\begin{equation}
\label{eq:wave_amplitude_modulation}
\dfrac{\widetilde{a}}{a} = \sqrt{
  \dfrac{g}{\widetilde{g}}
  \dfrac{\widetilde{\sigma}}{\sigma}
  \dfrac{\widetilde{N}}{N}}
=
  \left( \dfrac{g}{\widetilde{g}} \right)^{0.25}
  \left( \dfrac{\widetilde{k}}{k} \right)^{0.25}
  \left( \dfrac{\widetilde{N}}{N} \right)^{0.5}
\end{equation}
To the first order in $\varepsilon_L$, and using (\ref{eq:k_short_exact})
and (\ref{eq:wave_action_modulation}), the above simplifies to the result of
\citet{longuet1960changes}:

\begin{equation}
\label{eq:wave_amplitude_modulation_order1}
\dfrac{\widetilde{a}}{a} = 
  \left( \dfrac{g}{\widetilde{g}} \right)^{0.25}
  \left( 1 + \varepsilon_L \right)^{0.75}
  \approx \frac{(1 + \varepsilon_L)^{0.75}}{(1 - \varepsilon_L)^{0.25}}
  \approx 1 + \varepsilon_L
\end{equation}

@@

To quantify the contribution of the nonlinear terms in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}), and to evaluate the steadiness assumption of the
analytical solutions, we now proceed to numerically integrate the full set of
crest and action conservation equations.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are discretised using
second-order central finite difference in space and integrated in time using the
fourth-order Runge-Kutta method \citep{butcher1996runge}.
The space is divided into 100 grid points.
This is effectively the same equation set and numerical configuration as that of
\citet{peureux2021unsteady}, except that their long-wave orbital velocities are
evaluated at $z=0$ instead of $z=\eta_L$, thus neglecting the Stokes drift
induced by long waves on short-wave groups
\citep{stokes1847,van2018stokes,monismith2020stokes}.
Their gravity modulation also does not consider the nonlinear or the curvature
effects, however, this difference does not qualitatively affect the results.
Another difference is the choice of the numerical scheme for spatial differences,
which in their case is the more sophisticated MUSCL4 scheme \citep{kurganov2000new}.
Although the second-order central finite difference is not appropriate for many
numerical problems, in our case it is sufficient because it is conservative and
the fields that we compute derivatives of are smooth and continuous.
The maximum relative error of the centred finite difference in this model
is $\approx 0.066\%$.
The wavenumber and wave action are conservative over time within
$\mathcal{O}(10^{-7})$ and $\mathcal{O}(10^{-5})$ relative error, respectively.
We show in the next section that our numerical solutions for infinite long-wave
trains are qualitatively equivalent to those of \citet{peureux2021unsteady}.

@@

\backsection[Acknowledgements]{
Thanks to Nathan Laxague, Fabrice Ardhuin, Nick Pizzo, and Peisen Tan for the
helpful discussions that improved this paper.
}
